Storm surges in the Strait of Georgia simulated with a regional model

### 2015-01-08 AO_guides/manuscriptAO.tex
Modified paper size and "To appear" header
Susan's comments

\documentclass{tATO2e}

@@

\author{Nancy Soontiens$^{1}$$^{\ast}$\thanks{$^\ast$Corresponding author's email: nsoontie@eos.ubc.ca
\vspace{6pt}}, Susan Allen$^{1}$,  Doug Latornell$^{1}$,  Kate Le Sou\"{e}f$^{1}$, Idalia Machuca$^{1}$,\\ Jean-Philippe Paquin$^{2}$,
Youyu Lu$^{3}$, Keith Thompson$^{2}$, Vasily Korabel$^{2}$
\\\vspace{6pt}  $^{1}${\em{University of British Columbia}};
$^{2}${\em{Dalhousie University}};
$^{3}${\em{Bedford Institute of Oceanography}}\\\received{v1.2 released April 2014} }

@@

%storm surge modelling
Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} in a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters \citep{murty1995storm, danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting, bernier2007tide, bernier2010tide}, relatively little has been published on storm surge modelling on the west coast of Canada in recent years. However, a storm surge forecasting program is currently operating over a domain that covers the entire British Columbia coastline \citep{StormBC}. Further, a statistical overview of past surge events in British Columbia waters suggests an increase in extreme water level events during warm El Nino periods \citep{abeys2011extreme}.

@@

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. To compliment previous barotropic storm surge modelling in this region, this model employs a higher resolution and includes the effect of stratification and rivers. However, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Several simulations comparing the effects of different forcing conditions are also discussed. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study.

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait as shown in Figure \ref{fig:domain}. Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells. As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis.

The curvilinear orthogonal numerical grid has been divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels were stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10 m the grid was stretched to a maximum grid spacing of 27 m at the lowest layer. At the bottom boundary, partial $z$-levels were utilized in order to limit large changes in bathymetry across grid cells \citep{madec2012nemo}.

@@

\begin{equation}
D = [(A_0 \cos g_0 - A_m \cos g_m)^2 + (A_0 \sin g_0 - A_m \sin g_m)^2]^{1/2}
\end{equation}\label{eq:compdiff}
where $A_0$, $A_m$, $g_0$ and $g_m$ are the observed and modelled amplitudes and phases. Figure \ref{fig:tides} compares modelled and observed amplitude and phase for the $M_2$ and $K_1$ tidal constituents for stations along a transect from the Strait of Juan de Fuca through to the Strait of Georgia and Johnstone Strait. Complex differences for these two constituents are also plotted.  Comparisons have been made with data from \citet{foreman1995tidal}, \citet{foreman2004m} and \citet{foreman2012circulation}. Station names and numbers are listed in Table  \ref{tab:comparison} in Appendix \ref{sec:appendix}.

@@

First, a comparison of the wind speed and direction for the two models and observations is provided in Figure \ref{fig:dec2012_weather} at Point Atkinson and at Sandheads, a lighthouse station in the southern Strait of Georgia. For each of the models, the wind speed and direction at the closest grid point to the longitude/latitude of Point Atkinson and Sandheads is used for comparison. Although the low resolution CGRF model has a grid spacing of 33 km, two grid points align closely with these locations. At Point Atkinson, the high resolution model is in very good agreement with observations for both the wind speed and direction, however, the low resolution CGRF model significantly underestimates the wind speed during high wind periods on Dec 15, 2012 and Dec 17, 2012. At Sandheads, both the low resolution and high resolution models have good agreement with observations. The low resolution model has a very good match at Sandheads likely because this grid cell happens to be mostly over water. Although not shown here, both models have a similar sea level atmospheric pressure at these two locations. Note that the HRDPS model does not employ data assimilation.

@@

\section{Conclusion and discussion}\label{sec:diss}
In this article, we have configured a three-dimensional numerical ocean model of the Salish Sea for accurate hindcasts of storm surges in the Strait of Georgia. The model is forced with eight tidal constituents at the western and northern open boundaries, as well as sea surface height anomaly from de-tided observations at Tofino and Port Hardy. Additionally, wind stress and local atmospheric pressure forcing were used at the ocean surface using an atmospheric reforecasting product called CGRF \citep{smith2014new}. Storm surge simulations began with zero initial conditions for the currents and sea surface height and were commenced several days prior to the event of interest to spin up the velocity fields.

Hindcasts of five significant events in February 2006, November 2006, December 2006, November 2009, and December 2012 were simulated. In general, the model performs well at reproducing the observed water level heights and residuals, with Willmott skill scores greater than 0.88 in all cases and generally lower skill scores for the residuals.  Further, in a single simulation, the skill scores are fairly consistent between the four stations examined. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean, or remote forcing,  contributes most significantly to the surge amplitude, in agreement with previous studies in this region \citep{murty1995storm}. Although the large scale wind patterns and atmospheric conditions are important in setting up the surge on the Pacific Ocean boundaries, the local winds within the Strait of Georgia are not the leading order contribution to surge amplitude. However, the local wind patterns can set up small spatial differences in the surge, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are typically higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions or a partial reflection from the mainland coast. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important north the San Juan and Gulf Islands. The modelled surge amplitude increased by about 10\% at Point Atkinson when the tide-surge interaction was neglected in one example.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management. Further, the remote forcing is applied uniformly across the open boundary and so any spatial variability in the incoming surge are not accounted for in the model. Since the remote forcing is an important contributing factor, any forecasting with this regional model must make use of results from a larger scale prediction system. Finally, stratification effects on storm surge will be considered in a future study to determine if a barotropic model is reasonably accurate.

>>>>

\documentclass[letterpaper]{tATO2e}

@@

\author{Nancy Soontiens$^{1}$$^{\ast}$\thanks{$^\ast$Corresponding author's email: nsoontie@eos.ubc.ca
\vspace{6pt}}, Susan E. Allen$^{1}$,  Doug Latornell$^{1}$,  Kate Le Sou\"{e}f$^{1}$, Idalia Machuca$^{1}$,\\ Jean-Philippe Paquin$^{2}$,
Youyu Lu$^{3}$, Keith Thompson$^{2}$, Vasily Korabel$^{2}$
\\\vspace{6pt}  $^{1}${\em{University of British Columbia}};
$^{2}${\em{Dalhousie University}};
$^{3}${\em{Bedford Institute of Oceanography}}\\\received{v1.2 released April 2014} }

@@

%storm surge modelling
Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} using a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters \citep{murty1995storm, danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting, bernier2007tide, bernier2010tide}, relatively little has been published on storm surge modelling on the west coast of Canada in recent years. However, a storm surge forecasting program is currently operating over a domain that covers the entire British Columbia coastline \citep{StormBC}. Further, a statistical overview of past surge events in British Columbia waters suggests an increase in extreme water level events during warm El Nino periods \citep{abeys2011extreme}.

@@

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. To compliment previous barotropic storm surge modelling in this region, this model employs a higher resolution and includes the effect of stratification and rivers. However, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is assessed in section \ref{sec:storm}. Several simulations comparing the effects of different forcing conditions are also discussed. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study.

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait (Figure \ref{fig:domain}). Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells. As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis.

The curvilinear orthogonal numerical grid has been divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels were stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10 m the grid was stretched to a maximum grid spacing of 27 m in the lowest layer. At the bottom boundary, partial $z$-levels were utilized in order to limit large changes in bathymetry across grid cells \citep{madec2012nemo}.

@@

\begin{equation}
D = [(A_0 \cos g_0 - A_m \cos g_m)^2 + (A_0 \sin g_0 - A_m \sin g_m)^2]^{1/2}
\end{equation}\label{eq:compdiff}
where $A_0$, $A_m$, $g_0$ and $g_m$ are the observed and modelled amplitudes and phases. Figure \ref{fig:tides} compares modelled and observed amplitude, phase, and complex differences for the $M_2$ and $K_1$ tidal constituents for stations along a transect from the Strait of Juan de Fuca through to the Strait of Georgia and Johnstone Strait. Comparisons have been made with data from \citet{foreman1995tidal}, \citet{foreman2004m} and \citet{foreman2012circulation}. Station names and numbers are listed in Table  \ref{tab:comparison} in Appendix \ref{sec:appendix}.

@@

First, a comparison of the wind speed and direction for the two models and observations is provided in Figure \ref{fig:dec2012_weather} at Point Atkinson and at Sandheads, a lighthouse station in the southern Strait of Georgia. For each of the models, the wind speed and direction at the closest grid point to the longitude/latitude of Point Atkinson and Sandheads is used for comparison. Although the low resolution CGRF model has a grid spacing of 33 km, two grid points align closely with these locations. At Point Atkinson, the high resolution model is in very good agreement with observations for both the wind speed and direction, however, the low resolution CGRF model significantly underestimates the wind speed during high wind periods on Dec 15, 2012 and Dec 17, 2012. At Sandheads, both the low resolution and high resolution models have good agreement with observations. The low resolution model has a very good match at Sandheads likely because this grid cell happens to be mostly over water. Although not shown here, both models have a similar sea level atmospheric pressure at these two locations. Note that data assimilation is used in the parent model of the nested HRDPS product.

@@

\section{Conclusion and discussion}\label{sec:diss}
In this article, we describe a three-dimensional numerical ocean model of the Salish Sea for accurate hindcasts of storm surges in the Strait of Georgia. The model is forced with eight tidal constituents at the western and northern open boundaries, as well as sea surface height anomaly from de-tided observations at Tofino and Port Hardy. Additionally, wind stress and local atmospheric pressure forcing were used at the ocean surface using an atmospheric reforecasting product called CGRF \citep{smith2014new}. Storm surge simulations began with zero initial conditions for the currents and sea surface height and were commenced several days prior to the event of interest to spin up the velocity fields.

Hindcasts of five significant events in February 2006, November 2006, December 2006, November 2009, and December 2012 were simulated. In general, the model performs well at reproducing the observed water level heights and residuals, with Willmott skill scores greater than 0.88 in all cases and generally higher skill scores for the water level heights.  Further, in a single simulation, the skill scores are fairly consistent between the four stations examined. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean, or remote forcing,  contributes most significantly to the surge amplitude, in agreement with previous studies in this region \citep{murty1995storm}. Although the large scale wind patterns and atmospheric conditions are important in setting up the surge on the Pacific Ocean boundaries, the local winds within the Strait of Georgia are not the leading order contribution to surge amplitude. However, the local wind patterns can set up small spatial differences in the surge, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are typically higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions or a partial reflection from the mainland coast. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important north the San Juan and Gulf Islands. The modelled surge amplitude increased by about 10\% at Point Atkinson when the tide-surge interaction was neglected in one example.

@@

Several factors considered in other storm surge modelling studies have been neglected in these simulations. Primarily, effects due to wetting and drying have not been included. This is an important factor to consider in shallow, low lying areas where water levels retreat significantly to reveal land. In this article, the four locations used to assess storm surge accuracy are not in shallow regions and so wetting and drying is not considered to be important. However, several communities in the Strait of Georgia, such as Boundary Bay and Crescent Beach, are in shallow regions and are typically most affected by storm surge during the winter season. As such, efforts to include wetting and drying will be considered in the future. Additionally, flooding effects due to wave run up is left for a more detailed inundation study with higher resolution in the near-shore areas. Local winds have a significant impact on wave action which is a primary concern for flooding management. Further, the remote forcing is applied uniformly across the open boundary and so any spatial variability in the incoming surge are not accounted for in the model. Since the remote forcing is an important contributing factor, any forecasting with this regional model must make use of results from a larger scale prediction system.
